Parsing Postfix log files

### 2007-05-30 doc/logparser.tex
Corrected and improved flow chart. Added graphs, hopefully clarified graphs too. Improved paper with clearer wording. Described field usage in example rule. Expanded Makefile to cover new graphs. Stupidly deleted everything, had to redo the days work.

        It is hoped that this will prove to have greater effectiveness as
        the number of rules grows; currently the effect is negligible
        in comparison to caching each compiled regex, but it still provides
        approximately a 10\% performance increase.

    \item [priority] This is the user-configurable companion to
        rule\_order: rules with a higher priority will be tried first,
        overriding rule\_order, allowing more specific rules to take
        precedence over more general rules.

@@

XXX CARL SUGGESTS EXPANDING ON THIS SOME MORE, BUT I DON'T KNOW HOW\@.
PERHAPS WHEN THE ALGORITHM IS PROPERLY EXPLAINED IT'LL BE EASIER\@?  MAKE
REFERENCES TO THE ALGORITHM, EXPLAIN MORE CLEARLY HOW THE FIELDS ARE USED,
HOW THE RULE MATCHES, AND SO ON\@.  SAVE\_BY\_PID IS GONE NOW, USE A
DIFFERENT EXAMPLE.

@@

name                & Unknown sender domain                             \\
description         & We do not accept mail from unknown domains        \\
program             & postfix/smtpd                                     \\
restriction\_name   & reject\_unknown\_sender\_domain                   \\
regex               & \verb!^<(__SENDER__)>: Sender address rejected: ! \\
                    & \verb!  Domain not found; from=<\1> !             \\
                    & \verb!  to=<(__RECIPIENT__)> proto=E?SMTP !       \\
                    & \verb!  helo=<(__HELO__)>$!                       \\
result\_cols        & recipient = 2; sender = 1                         \\
connection\_cols    & helo = 3                                          \\
result\_data        &                                                   \\
connection\_data    &                                                   \\
postfix\_action     & REJECTED                                          \\
action              & SAVE\_BY\_PID                                     \\
queueid             & 0                                                 \\
rule\_order         & 0                                                 \\
priority            & 0                                                 \\

@@

@@

        Process scheduling can have a still more confusing effect: quite
        often the child mail will be created, delivered and entirely
        finished with \textbf{before} the parent logs the creation line!
        Thus mails flagged as coming from an unknown origin cannot be
        entered into the database when their final log line is parsed;
        instead they must be marked as database ready and subsequently
        entered by the parent mail once it has been identified.  Quite
        apart from identifying mail injected in an unknown fashion, or bugs
        in the parser, unknown origin mails need to be marked as such
        because they lack some data present in the parent mail, and must
        copy that data from their parent before being entered in the
        database.  XXX EITHER REMOVE THE LAST LINE OR EXPAND ON IT SOME
        MORE\@.  FIGURE OUT WHY MORE DATA ISN'T REQUIRED\@.

@@

The intermingling of log entries from different mails immediately rules out
the possibility of handling each mail in isolation; the parser must be
capable of handling multiple mails in parallel, each potentially at a
different stage in its journey, without any interference between mails ---
except in the minority of cases where intra-mail interference is required.
I felt that the best way to implement this was to maintain state
information for every unfinished mail and manipulate the appropriate mail
correctly for each log line encountered.  The parser thus requires both a
method of mapping log lines to the correct mail and a method of specifying
the action the log line represents.  The former is achieved by using the
pid of the smtpd to identify the correct mail during the initial phase,
then switching to the queueid once the mail has been accepted.  The latter
uses the action field of the rule which matched the log line, executing the
code in the function named by the action.  From a high level viewpoint this
design shares significant similarity with how an editor maintains state for
multiple files and performs the actions the user initiates on the correct
file.

@@

The most likely cause of mis-identification is if a mail generated
internally by Postfix is identified as a bounce notification when it is a
different type of message; arguably this is a benefit rather than a
drawback, as future mails generated internally by Postfix will be handled
correctly.  XXX FIND OUT IF DSNS FIT THE PATTERN.

@@

@@

>>>>

        It is hoped that this will prove to have greater effectiveness as
        the number of rules grows; currently the effect is negligible
        in comparison to caching each compiled regex, but it still provides
        approximately a 7\% -- 10\% performance increase.

    \item [priority] This is the user-configurable companion to
        rule\_order: rules with a higher priority will be tried first,
        overriding rule\_order.  This allows more specific rules to take
        precedence over more general rules.

@@

@@

\textbf{Field}      & \textbf{Value}                                    \\
name                & Unknown sender domain                             \\
description         & We do not accept mail from unknown domains        \\
restriction\_name   & reject\_unknown\_sender\_domain                   \\
postfix\_action     & REJECTED                                          \\
program             & postfix/smtpd                                     \\
regex               & \verb!^__RESTRICTION_START__ <(__SENDER__)>: !    \\
                    & \verb!Sender address rejected: Domain not found;! \\
                    & \verb!from=<\5> to=<(__RECIPIENT__)> !            \\
                    & \verb!proto=E?SMTP helo=<(__HELO__)>$!            \\
result\_cols        & recipient = 6; sender = 5                         \\
connection\_cols    & helo = 7                                          \\
result\_data        &                                                   \\
connection\_data    &                                                   \\
action              & REJECTION                                         \\
queueid             & 1                                                 \\
rule\_order         & 0                                                 \\
priority            & 0                                                 \\

@@

The various fields are used as follows;

    \item [name, description, restriction\_name and postfix\_action:] are
        not \newline used by the algorithm, they serve to document the rule.

    \item [program and regex:] If the program equals the program which
        logged the line the regex will be matched against the line; if
        successful the action will be called, if not the next rule will be
        tried.

    \item [action:] will be called if the regex matches successfully (see
        Section \ref{actions-in-detail} for full details).

    \item [result\_cols, connection\_cols, result\_data and
        connection\_data:] are \newline used by the action to extract and
        save data matched by the regex.

    \item [queueid:] is used the extract the queueid matched by the regex,
        so that mails can be found by queueid and actions performed on the
        mail.

@@

        Process scheduling can have a still more confusing effect: quite
        often the child mail will be created, delivered and entirely
        finished with \textbf{before} the parent logs the creation line!
        Thus mails flagged as coming from an unknown origin cannot be
        entered into the database when their final log line is parsed;
        instead they must be marked as database ready and subsequently
        entered by the parent mail once it has been identified.  Quite
        apart from identifying mail injected in an unknown fashion, or bugs
        in the parser, unknown origin mails need to be marked as such
        because they lack some data present in the parent mail; they must
        copy that data from their parent before being entered in the
        database.

@@

The intermingling of log entries from different mails immediately rules out
the possibility of handling each mail in isolation; the parser must be
capable of handling multiple mails in parallel, each potentially at a
different stage in its journey, without any interference between mails ---
except in the minority of cases where intra-mail interference is required.
The best way to implement this is to maintain state information for every
unfinished mail and manipulate the appropriate mail correctly for each log
line encountered.  The parser thus requires both a method of mapping log
lines to the correct mail and a method of specifying the action the log
line represents.  The former is achieved by using the smtpd process id to
identify the correct mail during the initial phase, then switching to the
queueid once the mail has been accepted.  The latter uses the action field
of the rule which matched the log line, executing the code in the function
named by the action.  From a high level viewpoint this design shares
significant similarity with how an editor maintains state for multiple
files and performs the actions the user initiates on the correct file.

@@

The most likely cause of mis-identification is if a mail generated
internally by Postfix is identified as a bounce notification when it is a
different type of message; arguably this is a benefit rather than a
drawback, as future mails generated internally by Postfix will be handled
correctly.

@@

XXX CHANGE REFERENCES TO BIBLIOGRAPHY??  MERGE PREVIOUS SECTIONS??  PUT
BIBLIOGRAPHY IN ITS OWN SECTION??

@@

\subsection{execution time vs file size vs number lines}
\label{execution time vs file size vs number lines}
\includegraphics{plot-normal-filesize-numlines.ps}

\subsection{execution time vs file size vs number lines factor}
\label{execution time vs file size vs number lines factor}
\includegraphics{plot-normal-filesize-numlines-factor.ps}
